---
title: "Bordeaux 2019 - robust stats - part 6: correlation"
author: "Guillaume A. Rousselet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
---

```{r message=FALSE, warning=FALSE}
# dependencies
library(ggplot2)
library(tibble)
source("./code/Rallfun-v35.txt")
```

```{r}
sessionInfo()
```

# Generate data

In this example we sample from 2 uncorrelated variables. By chance there seems to be a non negligeable correlation. Changing the random seed or commenting out the line `set.seed(21)` will give different results. You can also sample trials from variables with a true correlation by changing `rho`.

```{r}
set.seed(21)
n <- 50 # sample size
mu <- c(0, 0) # means of the variables
rho <- 0 # correlation between variables
sigma <- matrix(c(1, rho, rho, 1), nrow = 2, byrow = TRUE) # covariance matrix
data <- MASS::mvrnorm(n = n, mu = mu, Sigma = sigma)
x <- data[,1]
y <- data[,2]

# make data frame
df <- tibble(x = x,
             y = y)

# ggplot figure
ggplot(df, aes(x = x, y = y)) + theme_classic() +
  # geom_hline(yintercept = 0) +
  # geom_vline(xintercept = 0) +
  geom_point(alpha = 0.6, size = 3) +
  geom_smooth(method='lm',formula=y~x) +
  theme(axis.title = element_text(size = 15, colour = "black"),
        axis.text = element_text(size = 13, colour = "black"),
        strip.text = element_text(size = 15, face = "bold")) +
  # scale_x_continuous(limits = c(-4, 4),
  #                    breaks = seq(-4, 4, 1)) +
  labs(x = expression(italic("Variable A")), y = expression(italic("Variable B")))
# ggsave(filename = "./corr_samp.pdf")
```

# Pearson's correlation
```{r}
out <- cor.test(x,y, method = "pearson")
out
```

# Percentile bootstrap confidence interval

## Pearson correlation
```{r}
pcorb(x,y, SEED = FALSE)
```

### Pearson correlation: detailed code
```{r}
set.seed(21)
nboot <- 599
# sample pairs of observations with replacement
data <- matrix(sample(length(y),size=length(y)*nboot,replace=TRUE),nrow=nboot)
# compute correlation for each pair
bvec <- apply(data,1,pcorbsub,x,y) # A 1 by nboot matrix.
# confidence interval is computed using special adjustments to account for heteroscedasticity
ilow<-15
ihi<-584
if(length(y) < 250){
ilow<-14
ihi<-585
}
if(length(y) < 180){
ilow<-11
ihi<-588
}
if(length(y) < 80){
ilow<-8
ihi<-592
}
if(length(y) < 40){
ilow<-7
ihi<-593
}
bsort <- sort(bvec)
ci <-c (bsort[ilow],bsort[ihi])

ggplot(as_tibble(bvec), aes(x = value)) + theme_bw() +
  geom_histogram(aes(y = ..density..), bins = 50,
                 fill = "white", colour = "black") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16)) +
  labs(x = "Bootstrap correlations") 
# ggsave(filename = "./pboot_dist.pdf")
```

## percentage bend correlation
```{r}
corb(x,y, corfun = pbcor, SEED = FALSE)
```

## 25% Winsorized correlation
```{r}
corb(x,y, corfun=wincor, tr=0.25, SEED = FALSE)
```

## skipped correlation: Pearson
```{r}
mscor(cbind(x,y),corfun=pcor)
```

## skipped correlation: Spearman
```{r}
mscor(cbind(x,y),corfun=spear)
```

# Compare correlations

The design of the study presented in Davis et al. (2008) is a mix between the independent and dependent cases: there are 2 independent groups, and for each group there are 3 dependent measurements. To build confidence intervals for the differences of differences of correlations, you would need to write your own code. Let's look at the simpler cases.

## Independent case

Generate data
```{r}
set.seed(21)
n <- 50 # sample size
mu <- c(0, 0) # means of the variables
rho <- 0 # correlation between variables
sigma <- matrix(c(1, rho, rho, 1), nrow = 2, byrow = TRUE) # covariance matrix

# group 1
data <- MASS::mvrnorm(n = n, mu = mu, Sigma = sigma)
x1 <- data[,1]
y1 <- data[,2]

# group 2
data <- MASS::mvrnorm(n = n, mu = mu, Sigma = sigma)
x2 <- data[,1]
y2 <- data[,2]

# make data frame
df <- tibble(x = c(x1, x2),
             y = c(y1, y2),
             group = factor(c(rep("group1",n),rep("group2",n))))

# ggplot figure
ggplot(df, aes(x = x, y = y)) + theme_classic() +
  # geom_hline(yintercept = 0) +
  # geom_vline(xintercept = 0) +
  geom_point(alpha = 0.6, size = 3) +
  geom_smooth(method='lm',formula=y~x) +
  theme(axis.title = element_text(size = 15, colour = "black"),
        axis.text = element_text(size = 13, colour = "black"),
        strip.text = element_text(size = 15, face = "bold")) +
  # scale_x_continuous(limits = c(-4, 4),
  #                    breaks = seq(-4, 4, 1)) +
  labs(x = expression(italic("Variable A")), y = expression(italic("Variable B"))) +
  facet_grid(cols = vars(group))
 # ggsave(filename = "./corr_samp_ind.pdf")
```


### Two Pearson correlations
```{r}
twopcor(x1,y1,x2,y2, SEED = FALSE)
```

### Two robust correlations
```{r}
twocor(x1,y1,x2,y2, corfun = pbcor)
```

## Dependent case

### Overlapping case
For instance, if we have 3 dependent variables, we want to compare the correlation between 1 and 3 to the correlation between 2 and 3.

Generate data
```{r}
set.seed(21)
n <- 50 # sample size
mu <- c(0, 0, 0) # means of the variables
rho12 <- 0.8 # correlation between variables 1 and 2
rho13 <- 0.2 # correlation between variables 1 and 3
rho23 <- 0.6 # correlation between variables 2 and 3
# define covariance matrix
sigma <- matrix(c(1, rho12, rho13, 
                  rho12, 1, rho23,
                  rho13, rho23, 1), 
                nrow = 3, byrow = TRUE) 

data <- MASS::mvrnorm(n = n, mu = mu, Sigma = sigma)
x <- data[,1:2]
y <- data[,3]

# make data frame
df <- tibble(x = x[,1],
             y = y)
# ggplot figure
ggplot(df, aes(x = x, y = y)) + theme_classic() +
  geom_point(alpha = 0.6, size = 3) +
  geom_smooth(method='lm',formula=y~x) +
  theme(axis.title = element_text(size = 15, colour = "black"),
        axis.text = element_text(size = 13, colour = "black"),
        strip.text = element_text(size = 15, face = "bold")) +
  # scale_x_continuous(limits = c(-4, 4),
  #                    breaks = seq(-4, 4, 1)) +
  labs(x = expression(italic("Variable A")), y = expression(italic("Variable C")))
 # ggsave(filename = "./corr_samp_dep_ov1.pdf")
 
 # make data frame
df <- tibble(x = x[,2],
             y = y)
# ggplot figure
ggplot(df, aes(x = x, y = y)) + theme_classic() +
  geom_point(alpha = 0.6, size = 3) +
  geom_smooth(method='lm',formula=y~x) +
  theme(axis.title = element_text(size = 15, colour = "black"),
        axis.text = element_text(size = 13, colour = "black"),
        strip.text = element_text(size = 15, face = "bold")) +
  # scale_x_continuous(limits = c(-4, 4),
  #                    breaks = seq(-4, 4, 1)) +
  labs(x = expression(italic("Variable B")), y = expression(italic("Variable C")))
 # ggsave(filename = "./corr_samp_dep_ov2.pdf")
```

For Pearson correlation:
```{r}
# x = matrix with 2 columns
# y = vector
TWOpov(x,y)
#TWOpovPV to get a p-value
```

For a robust correlation:
```{r}
twoDcorR(x,y, corfun=wincor, SEED=FALSE)
```

### Non-overlapping case
For instance, if we have 4 dependent variables, we want to compare the correlation between 1 and 2 to the correlation between 3 and 4.

For Pearson correlation:
```{r}
# x = matrix with 2 columns
# y = matrix with 2 columns
TWOpNOV(x,y)
```

For a robust correlation:
```{r}
twoDNOV(x,y, corfun=wincor, SEED=FALSE)
```
