---
title: "Bordeaux 2019 - robust stats - part 5: percentile-t bootstrap"
author: "Guillaume A. Rousselet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
---

```{r message=FALSE, warning=FALSE}
# dependencies
library(ggplot2)
library(tibble)
```

```{r}
sessionInfo()
```

# Functions from Rand Wilcox

## One-sample
```{r, eval = FALSE}
# source('./code/Rallfun-v35.txt')
trimcibt(x, tr=0.2, alpha=0.1, nboot=1000, side=FALSE)
# tr = amount of trimming
# side = FALSE for an equal-tailed confidence interval
# side = TRUE for a symmetric confidence interval
```

## Two independent groups

That's the one you could try on your data.
```{r, eval = FALSE}
yuenbt(x,y,tr=0.2,alpha=0.05,nboot=599,side=FALSE)
```

## Two dependent groups
```{r, eval = FALSE}
ydbt(x,y,tr=0.2,alpha=0.05,nboot=599,side=FALSE,plotit=TRUE,op=1)
```

# Expected vs. empirical t distributions

```{r}
# population
m <- 50
s <- 20
location <- log(m^2 / sqrt(s^2 + m^2))
shape <- sqrt(log(1 + (s^2 / m^2)))

ggplot(data.frame(x = c(0, 130)), aes(x)) + theme_linedraw() +
  stat_function(fun = dlnorm, args = list(meanlog = location, sdlog = shape)) + 
labs(y = "Density") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size=20)) +
  ggtitle(paste0("Log normal distribution: mean = ",m,", sd = ",s))
 # ggsave(filename = './lnorm_dist.pdf')
```

```{r}
# sampling distribution
set.seed(111) 
n <- 20
nsim <- 10000
samp.dist <- apply(matrix(rlnorm(n*nsim, location, shape),nrow = nsim), 1, mean)

ggplot(data.frame(x=samp.dist), aes(x)) + theme_linedraw() +
  geom_histogram(aes(y = ..density..), colour = "black", fill = "white", bins = 50) +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size=20)) +
  ggtitle(paste(nsim,"sample means: mean =",round(mean(samp.dist),digits=1),", sd = ",round(sd(samp.dist),digits=2)))
# ggsave(filename = './lnorm_sampdist.pdf')
```

```{r}
# one sample
set.seed(4)
n <- 20
samp <- rlnorm(n, location, shape)
t.test(samp)
```

T-value
```{r}
# Formula
tval <- function(x,nv){
  tval <- (mean(x) - nv) / sqrt(var(x)/length(x))
  tval
}
```

Theoretical t distribution
```{r}
df <- n-1

ggplot(data.frame(x = c(-5, 5)), aes(x)) + theme_linedraw() +
  stat_function(fun = dt, args = list(df = df)) + 
labs(y = "Density") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size=20)) +
  ggtitle(paste0("t distribution: df = ",df))
# ggsave(filename = './theor_t.pdf')
```

```{r}
# expected quantile to use in CI formula
alpha <- 0.05
qt.th<- qt(1-alpha/2, df)
qt.th
```


```{r}
# sampling distribution of t
set.seed(111) 
n <- 20
nsim <- 10000
samp.dist <- apply(matrix(rlnorm(n*nsim, location, shape),nrow = nsim), 1, tval, nv = m)

ggplot(data.frame(x=samp.dist), aes(x)) + theme_linedraw() +
  geom_histogram(aes(y=..density..), 
                 colour = "black", fill = "white", bins = 50) +
  # stat_function(fun = dt, args = list(df = df)) + 
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size=20)) +
  ggtitle(paste(nsim,"sample t values: mean =",round(mean(samp.dist),digits=1)))
# ggsave(filename = './sampdist_t.pdf')
```

```{r}
# sampling distribution of t + theoretical
set.seed(111) 
n <- 20
nsim <- 10000
samp.dist <- apply(matrix(rlnorm(n*nsim, location, shape),nrow = nsim), 1, tval, nv = m)

ggplot(data.frame(x=samp.dist), aes(x)) + theme_linedraw() +
  geom_histogram(aes(y=..density..), 
                 colour = "black", fill = "white", bins = 50) +
  stat_function(fun = dt, args = list(df = df), colour = "red") + 
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size=20)) +
  ggtitle(paste(nsim,"sample t values vs. empirical t values"))
# ggsave(filename = './comp_theor_samp_t.pdf')
```

```{r}
# quantiles based on sampling distribution
qt.emp <- quantile(samp.dist, c(alpha/2, 1-(alpha/2)))
```

```{r}
# sampling distribution of t + theoretical:
# quantile t values
set.seed(111) 
n <- 20
nsim <- 10000
samp.dist <- apply(matrix(rlnorm(n*nsim, location, shape),nrow = nsim), 1, tval, nv = m)

ggplot(data.frame(x=samp.dist), aes(x)) + theme_linedraw() +
  geom_histogram(aes(y=..density..), 
                 colour = "black", fill = "white", bins = 50) +
  geom_vline(xintercept = qt.emp, colour = "black", size = 1.5) +
  geom_vline(xintercept = c(-qt.th,qt.th) , colour = "red", size = 1.5) +
  stat_function(fun = dt, args = list(df = df), colour = "red") + 
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size=20)) +
  ggtitle(paste(nsim,"sample t values vs. empirical t values"))
# ggsave(filename = './comp_theor_samp_t_q.pdf')
```

## Different confidence intervals

Standard confidence interval using t value theoretical sampling distribution
```{r, fig.height=1}
set.seed(21)
n <- 20
samp <- rlnorm(n, location, shape)
ci.st <- t.test(samp)$conf.int
ggplot(data = data.frame(x = rep("gr", n), y = samp), aes(x=x,y=y)) +
  theme_linedraw() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank()) +
  geom_jitter(alpha = 0.5) +
  geom_hline(yintercept = mean(samp)) +
  geom_segment(y = ci.st[1], yend = ci.st[2], x = 1, xend = 1, colour = "red") +
  scale_y_continuous(breaks = seq(20, 100, 10)) +
  coord_flip()
# ggsave(filename = './ci1.pdf')
```

Check values
```{r}
ci.st
```

Check confidence interval by hand
```{r}
alpha <- 0.05
df <- n-1 # degrees of freedom
samp.m <- mean(samp)
sem <- sd(samp) / sqrt(n)
ci <- vector("numeric",2)
ci[1] <- samp.m - qt(1-alpha/2, df)  * sem
ci[2] <- samp.m + qt(1-alpha/2, df)  * sem
ci
```

Confidence interval using empiricial t value sampling distribution
```{r}
alpha <- 0.05
df <- n-1 # degrees of freedom
samp.m <- mean(samp)
sem <- sd(samp) / sqrt(n)
ci.emp <- vector("numeric",2)
ci.emp[1] <- samp.m - qt.emp[2]  * sem
ci.emp[2] <- samp.m - qt.emp[1]  * sem
ci.emp
```

Add CI using empirical t distribution
```{r, fig.height=1}
set.seed(21)
n <- 20
samp <- rlnorm(n, location, shape)
ci.st <- t.test(samp)$conf.int
ggplot(data = data.frame(x = rep("gr", n), y = samp), aes(x=x,y=y)) +
  theme_linedraw() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank()) +
  geom_jitter(alpha = 0.5) +
  geom_hline(yintercept = mean(samp)) +
  geom_segment(y = ci.st[1], yend = ci.st[2], x = 1, xend = 1, colour = "red") +
  geom_segment(y = ci.emp[1], yend = ci.emp[2], x = 1.02, xend = 1.02, colour = "black") +
  scale_y_continuous(breaks = seq(20, 100, 10)) +
  coord_flip()
# ggsave(filename = './ci2.pdf')
```

# Percentile t-bootstrap confidence interval
```{r}
set.seed(333)
out <- trimcibt(samp, tr=0, alpha=0.05, nboot=1000, side=FALSE, SEED = FALSE)
out
```

## Bootstrap estimate of the sampling distribution of t
```{r}
set.seed(333)
nboot <- 1000
tr <- 0
alpha <- 0.05 
ibot <- round(alpha*nboot/2)
itop <- nboot-ibot
sem <- sd(samp)/sqrt(n)
samp.m <- mean(samp, tr)
# mean centre the data
csamp <- samp - mean(samp, trim = tr)
# sample with replacement
bootdata <- matrix(sample(csamp,size=n*nboot,replace=TRUE),nrow=nboot)
boot.tval <- sort(apply(bootdata,1,mean,tr) / (apply(bootdata,1,sd,tr) / sqrt(n)))
ci.ptb <- samp.m - boot.tval[itop] * sem
ci.ptb[2] <- samp.m - boot.tval[ibot] * sem
ci.ptb
```

## Superimpose t distributions
Histogram (black) = sampling distribution
Red = theoretical
Green = bootstrap t distribution

```{r}
# sampling distribution of t + theoretical
set.seed(111) 
n <- 20
nsim <- 10000
samp.dist <- apply(matrix(rlnorm(n*nsim, location, shape),nrow = nsim), 1, tval, nv = m)

ggplot(data.frame(x=samp.dist), aes(x)) + theme_linedraw() +
  geom_histogram(aes(y=..density..), 
                 colour = "black", fill = "white", bins = 50) +
  stat_function(fun = dt, args = list(df = df), colour = "red", size = 1) + 
  geom_line(data=tibble(x = boot.tval), aes(x=x, y = ..density..), 
            stat = 'density', colour = "green3", size = 1) + 
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size=18)) +
  ggtitle(paste(nsim,"sample t values / theoretical t (red) / boot t (green)"))
ggsave(filename = './comp_theor_samp_boot_t.pdf')
```

## Add ptb confidence interval
```{r, fig.height=1}
set.seed(21)
n <- 20
samp <- rlnorm(n, location, shape)
ci.st <- t.test(samp)$conf.int
ggplot(data = data.frame(x = rep("gr", n), y = samp), aes(x=x,y=y)) +
  theme_linedraw() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank()) +
  geom_jitter(alpha = 0.5) +
  geom_hline(yintercept = mean(samp)) +
  geom_segment(y = ci.st[1], yend = ci.st[2], x = 1, xend = 1, colour = "red") +
  geom_segment(y = ci.emp[1], yend = ci.emp[2], x = 1.02, xend = 1.02, colour = "black") +
  geom_segment(y = out$ci[1], yend = out$ci[2], x = 0.98, xend = 0.98, colour = "green3") +
  scale_y_continuous(breaks = seq(20, 100, 10)) +
  coord_flip()
# ggsave(filename = './ci3.pdf')
```

