---
title: "Bordeaux 2019 - robust stats - part 4: percentile bootstrap - confidence interval variability"
author: "Guillaume A. Rousselet"
date: "`r Sys.Date()`"
output:
  github_document:
    html_preview: yes
    toc: yes
    toc_depth: 2
  # pdf_document:
  #   fig_caption: no
  #   number_sections: no
  #   toc: yes
  #   toc_depth: 2
---

```{r message=FALSE, warning=FALSE}
# dependencies
library(ggplot2)
library(tibble)
```

```{r}
sessionInfo()
```

This section is concerned with the long term performance of our methods, looking in particular at the behaviour of confidence intervals. The demonstrations and exercises proposed here could be expanded by considering the simulations and results from this article:
[What Teachers Should Know About the Bootstrap: Resampling in the Undergraduate Statistics Curriculum](https://amstat.tandfonline.com/doi/abs/10.1080/00031305.2015.1089789#.XJjIty2cbUI)

# Confidence interval variability

We perform many experiments, and for each experiment we compute a confidence interval, which we plot as horizontal lines.

## Sample from a normal distribution
```{r}
nboot <- 1000
alpha <- .1
lo <- nboot*(alpha/2)
hi <- nboot - lo
lo <- lo + 1

nexp <- 50 # number of experiments
n <- 20 # sample size in each experiment
n.m <- 50 # mean of normal distribution 
n.sd <- 20 # sd of normal distribution
ci <- matrix(NA, nrow = nexp, ncol = 2)
ci.ttest <- matrix(NA, nrow = nexp, ncol = 2)
for(E in 1:nexp){
  # sample data + bootstrap + compute mean of each bootstrap sample
  samp <- rnorm(n, n.m, n.sd)
  boot.samp <- apply(matrix(sample(samp, n*nboot, replace = TRUE), nrow = nboot), 1, mean)
  sort.boot.samp <- sort(boot.samp) # sort bootstrap estimates
  ci[E,1] <- sort.boot.samp[lo] 
  ci[E,2] <- sort.boot.samp[hi]
  ci.ttest[E,] <- t.test(samp, conf.level = 1-alpha)$conf.int
}
```

### Illustrate results: bootstrap
```{r, fig.height=3, fig.width=2}
df <- tibble(x = as.vector(ci),
             y = rep(1:nexp,2),
             gr = factor(rep(1:nexp,2))
            )
ggplot(df, aes(x = x, y = y)) + theme_linedraw() +
  geom_point() +
  geom_line(aes(group = gr)) +
  geom_vline(xintercept = n.m, colour = "red") +
  labs(x = "Values", y = "Experiments") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))

# https://stackoverflow.com/questions/12253239/vertical-lines-between-points-with-ggplot2
# ggsave(filename = './50expt_bootci.pdf')
```

### Illustrate results: t-test
```{r, fig.height=3, fig.width=2}
df <- tibble(x = as.vector(ci.ttest),
             y = rep(1:nexp,2),
             gr = factor(rep(1:nexp,2))
            )
ggplot(df, aes(x = x, y = y)) + theme_linedraw() +
  geom_point() +
  geom_line(aes(group = gr)) +
  geom_vline(xintercept = n.m, colour = "red") +
  labs(x = "Values", y = "Experiments") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
# ggsave(filename = './50expt_ttci.pdf')
```

### Illustrate results: bootstrap vs. t-test
```{r, fig.height=3, fig.width=2}

dof <- 0.1 # dodge factor
df <- tibble(x = c(as.vector(ci[,1]),as.vector(ci.ttest[,1])),
             xend = c(as.vector(ci[,2]),as.vector(ci.ttest[,2])),
             y = c((1:nexp)-dof,(1:nexp)+dof),
             yend = c((1:nexp)-dof,(1:nexp)+dof),
             gr = factor(rep(1:nexp,2)),
             ci = factor(c(rep('bootstrap',nexp), rep('ttest',nexp)))
            )

ggplot(df, aes(x = x, y = y, xend = xend, yend = yend, colour = ci, group = gr)) + 
  theme_linedraw() +
  geom_segment() +
  scale_size_manual(values = c(1,0.5)) +
  geom_vline(xintercept = n.m, colour = "red") +
  labs(x = "Values", y = "Experiments") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
# ggsave(filename = './50expt_compci.pdf')
```

### Width variability

Let's get more data
```{r}
set.seed(21)

nboot <- 200
alpha <- .1
lo <- nboot*(alpha/2)
hi <- nboot - lo
lo <- lo + 1

nexp <- 1000 # number of experiments
n <- 20 # sample size in each experiment
n.m <- 50 # mean of normal distribution 
n.sd <- 20 # sd of normal distribution
ci <- matrix(NA, nrow = nexp, ncol = 2)
ci.ttest <- matrix(NA, nrow = nexp, ncol = 2)
for(E in 1:nexp){
  # sample data + bootstrap + compute mean of each bootstrap sample 
  samp <- rnorm(n, n.m, n.sd)
  boot.samp <- apply(matrix(sample(samp, n*nboot, replace = TRUE), nrow = nboot), 1, mean)
  sort.boot.samp <- sort(boot.samp) # sort bootstrap estimates
  ci[E,1] <- sort.boot.samp[lo] 
  ci[E,2] <- sort.boot.samp[hi]
  ci.ttest[E,] <- t.test(samp, conf.level = 1-alpha)$conf.int
}
```

Illustrate distribution of the width of the confidence intervals
```{r}
df <- tibble(value = c(as.vector(ci[,2]-ci[,1]),
                       as.vector(ci.ttest[,2]-ci.ttest[,1])),
             ci = c(rep("bootstrap", nexp),rep("t-test", nexp))
             )
ggplot(df, aes(x = value, colour = ci)) +
  theme_linedraw() +
  geom_line(aes(y = ..density..), stat = 'density', size = 1) +
   theme(legend.position = c(0.1, 0.85),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 12),
      legend.title = element_text(size = 14)
       ) +
  scale_colour_manual(name = "Conf. int.", values = c("orange1", "blue")) +
  coord_cartesian(xlim = c(0, 35))
```

What happens if you change the sample size?
What happens if you decrease the number of boostrap samples? For instance, try nboot = 50.

### Coverage
How often does the confidence interval contain the population value?
Is it as the nominal (expected) value?
```{r}
mean(n.m > ci[,1] & n.m < ci[,2])
```

### Coverage of standard confidence interval
How often does the confidence interval contain the population value?
Is it as the nominal (expected) value?
```{r}
mean(n.m > ci.ttest[,1] & n.m < ci.ttest[,2])
```

## Sample from a normal distribution + outlier
```{r}
nboot <- 1000
alpha <- .1
lo <- nboot*(alpha/2)
hi <- nboot - lo
lo <- lo + 1

nexp <- 50 # number of experiments
n <- 20 # sample size in each experiment
n.m <- 50 # mean of normal distribution 
n.sd <- 20 # sd of normal distribution
ci <- matrix(NA, nrow = nexp, ncol = 2)
ci.ttest <- matrix(NA, nrow = nexp, ncol = 2)
ci.no <- matrix(NA, nrow = nexp, ncol = 2) # results without outlier
ci.ttest.no <- matrix(NA, nrow = nexp, ncol = 2) # results without outlier
for(E in 1:nexp){
  # sample data
  samp <- rnorm(n, n.m, n.sd)
  
  # bootstrap + compute mean of each bootstrap sample
  boot.samp <- apply(matrix(sample(samp, n*nboot, replace = TRUE), nrow = nboot), 1, mean)
  sort.boot.samp <- sort(boot.samp) # sort bootstrap estimates
  ci.no[E,1] <- sort.boot.samp[lo] 
  ci.no[E,2] <- sort.boot.samp[hi]
  ci.ttest.no[E,] <- t.test(samp, conf.level = 1-alpha)$conf.int
  
  # replace largest observation by outlier
  samp <- sort(samp)
  samp[n] <- samp[n]*2 
  # bootstrap + compute mean of each bootstrap sample
  boot.samp <- apply(matrix(sample(samp, n*nboot, replace = TRUE), nrow = nboot), 1, mean)
  sort.boot.samp <- sort(boot.samp) # sort bootstrap estimates
  ci[E,1] <- sort.boot.samp[lo] 
  ci[E,2] <- sort.boot.samp[hi]
  ci.ttest[E,] <- t.test(samp, conf.level = 1-alpha)$conf.int
}
```

### Illustrate results: bootstrap
```{r, fig.height=3, fig.width=2}
df <- tibble(x = as.vector(ci),
             y = rep(1:nexp,2),
             gr = factor(rep(1:nexp,2))
            )
ggplot(df, aes(x = x, y = y)) + theme_linedraw() +
  geom_point() +
  geom_line(aes(group = gr)) +
  geom_vline(xintercept = n.m, colour = "red") +
  labs(x = "Values", y = "Experiments") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
```

### Illustrate results: t-test
```{r, fig.height=3, fig.width=2}
df <- tibble(x = as.vector(ci.ttest),
             y = rep(1:nexp,2),
             gr = factor(rep(1:nexp,2))
            )
ggplot(df, aes(x = x, y = y)) + theme_linedraw() +
  geom_point() +
  geom_line(aes(group = gr)) +
  geom_vline(xintercept = n.m, colour = "red") +
  labs(x = "Values", y = "Experiments") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
```

### Illustrate results: bootstrap vs. t-test
```{r, fig.height=3, fig.width=2}

dof <- 0.1 # dodge factor
df <- tibble(x = c(as.vector(ci[,1]),as.vector(ci.ttest[,1])),
             xend = c(as.vector(ci[,2]),as.vector(ci.ttest[,2])),
             y = c((1:nexp)-dof,(1:nexp)+dof),
             yend = c((1:nexp)-dof,(1:nexp)+dof),
             gr = factor(rep(1:nexp,2)),
             ci = factor(c(rep('bootstrap',nexp), rep('ttest',nexp)))
            )

ggplot(df, aes(x = x, y = y, xend = xend, yend = yend, colour = ci, group = gr)) + 
  theme_linedraw() +
  geom_segment() +
  scale_size_manual(values = c(1,0.5)) +
  geom_vline(xintercept = n.m, colour = "red") +
  labs(x = "Values", y = "Experiments") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
```

### Illustrate results: t-test with and without outlier
```{r, fig.height=3, fig.width=2}

dof <- 0.1 # dodge factor
df <- tibble(x = c(as.vector(ci.ttest.no[,1]),as.vector(ci.ttest[,1])),
             xend = c(as.vector(ci.ttest.no[,2]),as.vector(ci.ttest[,2])),
             y = c((1:nexp)-dof,(1:nexp)+dof),
             yend = c((1:nexp)-dof,(1:nexp)+dof),
             gr = factor(rep(1:nexp,2)),
             ci = factor(c(rep('No outlier',nexp), rep('Outlier',nexp)))
            )

ggplot(df, aes(x = x, y = y, xend = xend, yend = yend, colour = ci, group = gr)) + 
  theme_linedraw() +
  geom_segment() +
  scale_size_manual(values = c(1,0.5)) +
  geom_vline(xintercept = n.m, colour = "red") +
  labs(x = "Values", y = "Experiments") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14)) +
  ggtitle("T-test confidence intervals")
```

### Illustrate results: bootstrap with and without outlier
```{r, fig.height=3, fig.width=2}

dof <- 0.1 # dodge factor
df <- tibble(x = c(as.vector(ci.no[,1]),as.vector(ci[,1])),
             xend = c(as.vector(ci.no[,2]),as.vector(ci[,2])),
             y = c((1:nexp)-dof,(1:nexp)+dof),
             yend = c((1:nexp)-dof,(1:nexp)+dof),
             gr = factor(rep(1:nexp,2)),
             ci = factor(c(rep('No outlier',nexp), rep('Outlier',nexp)))
            )

ggplot(df, aes(x = x, y = y, xend = xend, yend = yend, colour = ci, group = gr)) + 
  theme_linedraw() +
  geom_segment() +
  scale_size_manual(values = c(1,0.5)) +
  geom_vline(xintercept = n.m, colour = "red") +
  labs(x = "Values", y = "Experiments") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14)) +
  ggtitle("Bootstrap confidence intervals")
```

### Width variability: with outliers

Let's get more data
```{r}
set.seed(21)

nboot <- 200
alpha <- .1
lo <- nboot*(alpha/2)
hi <- nboot - lo
lo <- lo + 1

nexp <- 1000 # number of experiments
n <- 20 # sample size in each experiment
n.m <- 50 # mean of normal distribution 
n.sd <- 20 # sd of normal distribution
ci <- matrix(NA, nrow = nexp, ncol = 2)
ci.ttest <- matrix(NA, nrow = nexp, ncol = 2)
for(E in 1:nexp){
  # sample data
  samp <- rnorm(n, n.m, n.sd)
  # replace largest observation by outlier
  samp <- sort(samp)
  samp[n] <- samp[n]*2 
  # bootstrap + compute mean of each bootstrap sample 
  boot.samp <- apply(matrix(sample(samp, n*nboot, replace = TRUE), nrow = nboot), 1, mean)
  sort.boot.samp <- sort(boot.samp) # sort bootstrap estimates
  ci[E,1] <- sort.boot.samp[lo] 
  ci[E,2] <- sort.boot.samp[hi]
  ci.ttest[E,] <- t.test(samp, conf.level = 1-alpha)$conf.int
}
```

Illustrate distribution of the width of the confidence intervals

```{r}
df <- tibble(value = c(as.vector(ci[,2]-ci[,1]),
                       as.vector(ci.ttest[,2]-ci.ttest[,1])),
             ci = c(rep("bootstrap", nexp),rep("t-test", nexp))
             )
ggplot(df, aes(x = value, colour = ci)) +
  theme_linedraw() +
  geom_line(aes(y = ..density..), stat = 'density', size = 1) +
   theme(legend.position = c(0.1, 0.85),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 12),
      legend.title = element_text(size = 14)
       ) +
  scale_colour_manual(name = "Conf. int.", values = c("orange1", "blue")) +
  coord_cartesian(xlim = c(0, 35))
```

Scatterplot of width values
```{r}
df <- tibble(bootstrap = as.vector(ci[,2]-ci[,1]),
             ttest = as.vector(ci.ttest[,2]-ci.ttest[,1]))
ggplot(df, aes(x = ttest, y = bootstrap)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)
```
The t-test confidence intervals are systematically longer than the bootstrap ones.

### Coverage
How often does the confidence interval contain the population value?
Is it as the nominal (expected) value?
```{r}
mean(n.m > ci[,1] & n.m < ci[,2])
```

### Coverage of standard confidence interval
How often does the confidence interval contain the population value?
Is it as the nominal (expected) value?
```{r}
mean(n.m > ci.ttest[,1] & n.m < ci.ttest[,2])
```


## Sample from a gamma distribution + bootstrap CI

```{r}
nboot <- 1000
alpha <- .1
lo <- nboot*(alpha/2)
hi <- nboot - lo
lo <- lo + 1

nexp <- 50 # number of experiments
n <- 20 # sample size in each experiment

m <- 50 # mean of gamma distribution 
s <- 20 # sd of gamma distribution
location <- log(m^2 / sqrt(s^2 + m^2))
shape <- sqrt(log(1 + (s^2 / m^2)))

ci <- matrix(NA, nrow = nexp, ncol = 2)
ci.ttest <- matrix(NA, nrow = nexp, ncol = 2)

for(E in 1:nexp){
  # sample data + bootstrap + compute mean of each bootstrap sample
  samp <- rlnorm(n, meanlog = location, sdlog = shape)
  boot.samp <- apply(matrix(sample(samp, n*nboot, replace = TRUE), nrow = nboot), 1, mean)
  sort.boot.samp <- sort(boot.samp) # sort bootstrap estimates
  ci[E,1] <- sort.boot.samp[lo] 
  ci[E,2] <- sort.boot.samp[hi]
  ci.ttest[E,] <- t.test(samp, conf.level = 1-alpha)$conf.int
}
```

### Illustrate results: bootstrap CI
```{r, fig.height=3, fig.width=2}
df <- tibble(x = as.vector(ci),
             y = rep(1:nexp,2),
             gr = factor(rep(1:nexp,2))
            )
ggplot(df, aes(x = x, y = y)) + theme_linedraw() +
  geom_point() +
  geom_line(aes(group = gr)) +
  geom_vline(xintercept = m, colour = "red") +
  labs(x = "Values", y = "Experiments") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
```

### Illustrate results: standard CI
```{r, fig.height=3, fig.width=2}
df <- tibble(x = as.vector(ci.ttest),
             y = rep(1:nexp,2),
             gr = factor(rep(1:nexp,2))
            )
ggplot(df, aes(x = x, y = y)) + theme_linedraw() +
  geom_point() +
  geom_line(aes(group = gr)) +
  geom_vline(xintercept = m, colour = "red") +
  labs(x = "Values", y = "Experiments") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
```

### Width variability

Let's get more data
```{r}
set.seed(21)

nboot <- 200
alpha <- .1
lo <- nboot*(alpha/2)
hi <- nboot - lo
lo <- lo + 1

m <- 50 # mean of gamma distribution 
s <- 20 # sd of gamma distribution
location <- log(m^2 / sqrt(s^2 + m^2))
shape <- sqrt(log(1 + (s^2 / m^2)))

nexp <- 1000 # number of experiments
n <- 20 # sample size in each experiment
ci <- matrix(NA, nrow = nexp, ncol = 2)
ci.ttest <- matrix(NA, nrow = nexp, ncol = 2)
for(E in 1:nexp){
  # sample data + bootstrap + compute mean of each bootstrap sample
  samp <- rlnorm(n, meanlog = location, sdlog = shape)
  boot.samp <- apply(matrix(sample(samp, n*nboot, replace = TRUE), nrow = nboot), 1, mean)
  sort.boot.samp <- sort(boot.samp) # sort bootstrap estimates
  ci[E,1] <- sort.boot.samp[lo] 
  ci[E,2] <- sort.boot.samp[hi]
  ci.ttest[E,] <- t.test(samp, conf.level = 1-alpha)$conf.int
}
```

Illustrate distribution of the width of the confidence intervals

```{r}
df <- tibble(value = c(as.vector(ci[,2]-ci[,1]),
                       as.vector(ci.ttest[,2]-ci.ttest[,1])),
             ci = c(rep("bootstrap", nexp),rep("t-test", nexp))
             )
ggplot(df, aes(x = value, colour = ci)) +
  theme_linedraw() +
  geom_line(aes(y = ..density..), stat = 'density', size = 1) +
   theme(legend.position = c(0.1, 0.85),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 12),
      legend.title = element_text(size = 14)
       ) +
  scale_colour_manual(name = "Conf. int.", values = c("orange1", "blue")) +
  coord_cartesian(xlim = c(0, 35))
```

What happens if you change the sample size? Try n = 10 for instance.
What happens if you decrease the number of boostrap samples? For instance, try nboot = 50.

### Coverage
How often does the bootstrap confidence interval contain the population value?
```{r}
mean(m > ci[,1] & m < ci[,2])
```

How often does the standard confidence interval contain the population value?
```{r}
mean(m > ci.ttest[,1] & m < ci.ttest[,2])
```

# Take-home exercises

- Repeat for different nboot.
- Plot mean width/coverage as a function of nboot and sample size.
- look at confidence intervals for the trimmed mean and the median (check out Rand Wilcox's functions `trimci()` for trimmed means and `sintv2()` for medians).





