---
title: "Bordeaux 2019 - robust stats - part 2: robust estimation"
author: "Guillaume A. Rousselet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
---

```{r message=FALSE, warning=FALSE}
# dependencies
library(ggplot2)
library(tibble)
source('./code/Rallfun-v35.txt')
```

```{r}
sessionInfo()
```

Here we focus on simple examples: 
- one-sample
- two independent samples
- two dependent samples

For inferences on multiple quantiles, see these guides about the [shift function](https://github.com/GRousselet/rogme) and the [hierarchical shift function](https://github.com/GRousselet/rogme/blob/master/docs/hsf.md).

# Quantiles
```{r}
# check help, many techniques available
xf<-rnorm(100)
quantile(xf)
# Harrel-Davis estimator of the quantiles
hd(xf,q=.5) # Harrel-Davis estimate of the median
# ideal fourths
idealf(xf) # better 1st and 3rd quartiles?
```

# Discarding outliers 

Crazy example
```{r}
x <- c(2,2,3,3,3,4,4,4,100000,100000)
```

## Boxplot rule
outbox(x)
res<-outbox(x)
xclean<-x[res$keep]

# 2 SD away from the mean rule
k<-2
res<-(abs(x-mean(x))/sd(x))>k
x[res]

# MAD median rule
k<-2.24
(abs(x-median(x))/(mad(x)))>k
#out(x)

# Measures of central tendency 

## Mean vs. median
```{r}
x<-seq(0.05,4,0.05)
fx<-df(x,6,18) #fx<-dlnorm(x)
plot(x,fx,type="l")
n<-10000
samp<-rf(n,6,18)
mx<-mean(samp)
mdx<-median(samp)
arrows( mx,df(mx,6,18),mx,0,col='red')
arrows( mdx,df(mdx,6,18),mdx,0,col='green')
```

## Trimmed means

### Illustrate 20% trimming - Normal distribution
```{r}
tr<-.2
xv<-seq(-4,4,0.01)
yv<-dnorm(xv)
plot(xv,yv,type="l")
zval<-qnorm(tr, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
polygon(c(xv[xv<=zval],zval),c(yv[xv<=zval],yv[xv==-4]),col=5)
polygon(c(xv[xv>=-zval],-zval),c(yv[xv>=-zval],yv[xv==4]),col=5)
```

## Illustrate 20% trimming - F distribution
```{r}
tr<-.2
xv<-seq(0.01,4,0.01)
yv<-df(xv,6,18) #fx<-dlnorm(x)
plot(xv,yv,type="l")
zval<-qf(tr,6,18)
polygon(c(xv[xv<=zval],zval),c(yv[xv<=zval],yv[xv==0.01]),col=5)
zval<-qf(1-tr,6,18)
polygon(c(xv[xv>=zval],zval),c(yv[xv>=zval],yv[xv==4]),col=5)
```


## example: get trimmed means of all columns of 
```{r}
set.seed(21)
v <- matrix(rnorm(50), nrow = 10) # reshape into 10 x 5 matrix
apply(v, 2, mean, tr=.2) # compute trimmed mean for each column
```


# Winzorised mean
```{r, eval = FALSE}
winmean(x,tr=.2)
```

# Winzorised variance
```{r, eval = FALSE}
winvar(x,tr=.2)
```

# Standard error of the trimmed mean
```{r, eval = FALSE}
trimse(x,tr=.2)
```

# compare techniques
```{r}
self<-t(read.table("./data/self-awareness.csv",header=F))
stripchart(self, method="jitter", pch=1)
cbind(mean=round(mean(self),digits=1),median=median(self),tm=round(tmean(self,tr=.2),digits=1))
```


# Tests and confidence intervals for groups

## Independent groups

### Kolmogorov-Smirnov test

See explanation in this [blog post](https://garstats.wordpress.com/2016/05/02/robust-effect-sizes-for-2-independent-groups/).

## T-tests on means and trimmed means
```{r}
yuen(x,y,tr=0.2,alpha=0.05)
# set tr=0 for a t-test on means
# compare output to t.test
```

### Compare medians
```{r}
msmed(x,y,alpha=0.05)
```

## Dependent groups

### T-tests on trimmed means
```{r}
yuend(x,y,tr=0.2,alpha=0.05)
```

### Compare medians
```{r}
# use percentile bootstrap - see part 4
```

# Effect of outlier on mean and t-test
```{r}
set.seed(3) # set the seed of the random number generator for reproducibility
data <- rnorm(10, mean = 1, sd = 1) # get random sample from normal population

summary(data)

data2 <- c(data,5)
summary(data2)
x <- list("data"=data, "data2"=data2)
stripchart(x,method="jitter",pch=1)

t.test(data,mu=0)
t.test(data2,mu=0)
```

